---
title: "Empirical precision and detection limits"
subtitle: "Estimate the instrument precision from your own record and state the confidence level of the minimal detectable flux"
engine: knitr
editor: visual
---

The [minimal detectable flux](bestflux.qmd#minimal-detectable-flux) (MDF) of `goFlux` is built on the instrument precision `prec` and the measurement time:

$$MDF = \frac{prec}{t} \times \frac{V \cdot P}{R \cdot T \cdot A}$$

By default `prec` is the manufacturer's specification, given at import (e.g. `prec = c(3.5, 0.6, 45)` for the LI-7810). Field-worn analyzers are noisier than their datasheet, so an MDF built on the datasheet value is optimistic. Two things help: estimate the precision from the data, and say at which confidence level a flux counts as detected.

## Estimate the precision from the closures

`empirical.prec()` estimates the precision (1 σ) closure by closure from the **second differences** of the concentration inside the fitted window (rows with `flag == 1`):

$$\sigma_i = \frac{MAD(x_{k+1} - 2x_k + x_{k-1})}{\sqrt{6}}$$

A straight line, i.e. the flux, cancels exactly in a second difference, while white noise of standard deviation σ gives $Var = 6\sigma^2$. This is the one-sample Hadamard deviation, the drift-insensitive counterpart of the Allan deviation (Riley 2008, NIST SP 1065). MAD is the normal-consistent median absolute deviation (`stats::mad`, constant 1.4826). The precision of a group of closures (an analyzer on a field day, say) is the **median** of $\sigma_i$ over its closures, so a few large, curved or bubbly closures cannot set it.

```{r}
#| eval: false
library(goFlux)
data(manID.UGGA)

# median over the closures of the per-closure second-difference precision
empirical.prec(manID.UGGA, "CH4dry_ppb")

# one value per field day, campaign, analyzer ... (any column of the data frame)
empirical.prec(manID.UGGA, "CH4dry_ppb", by = "DATE")

# the per-closure values
attr(empirical.prec(manID.UGGA, "CH4dry_ppb"), "closures")

# one value per measurement, Allan deviation SD(dx)/sqrt(2) on the flagged window
empirical.prec(manID.UGGA, "CH4dry_ppb", method = "allan")
```

Second differences never span a gap or a change of logging interval; if the closures of a group were logged at different intervals, one row per interval (`dt_s`) is returned with a warning.

### The checks reported with σ

| column | expected | warns when | meaning |
|---|---|---|---|
| `ac1` | −2/3 | > −0.5 | lag-1 autocorrelation of the second differences: red or drift-dominated noise, which σ understates over a closure |
| `zero.frac` | ~0 | ≥ 0.3 | fraction of exactly-zero first differences: the analyzer logs one gas per row (e.g. Picarro G4301) or the record is quantized; keep each gas's fresh rows first |
| `d1c.ratio` | ~1 | outside 0.8–1.2 | σ from first differences centred on each closure's own median, over σ: trend or curvature leaking into first differences |

### Why not the whole-record first-difference MAD

Earlier versions of this function took the MAD of the *first* differences of the whole record. A first difference contains the closure's slope × Δt, and closures differ in slope, so the spread of slopes *between* closures is counted as noise (two noise-free closures rising 0 and 10 ppb per step give ~5 ppb). The effect is small at 1 Hz but reaches 1.4–2.3× at 5–10 s logging. That estimator is kept as `method = "mad_diff1"` for comparison with earlier results only.

### What σ cannot see: drift

Any difference-based precision is blind to instrument drift and to slow leaks: both change the concentration smoothly over a closure and so look like a flux, not like noise. Make periodic **blank closures** on an inert surface (a sealed chamber on a plate or foil); their apparent flux is the drift or leak floor. The trend of the ambient record around the closures can flag an unstable analyzer but cannot measure drift inside the chamber.

### Compare with the datasheet at your logging interval

Manufacturers quote precision at 1 s. A record logged every Δt seconds averages white noise down by $\sqrt{\Delta t}$, so compare the `prec` of a 5 s or 10 s record with `spec.at.interval(spec_1s, dt_s)` (the datasheet value ÷ $\sqrt{\Delta t}$), not with the 1 s figure; otherwise an analyzer can appear to beat its datasheet when it is several times worse.

## State the confidence level

With `prec` = 1 σ, `MDF = prec / t × flux.term` is a one-sigma limit. `MDF()`, `goFlux()` and `best.flux()` accept `conf`, a two-sided confidence level: the precision is multiplied by $z = qnorm(1 - (1 - conf)/2)$, i.e. 1.96 for `conf = 0.95` (the two-sigma criterion of Wassmann et al., 2018) and 1.645 for `conf = 0.90`. Without `conf` nothing changes ($z = 1$).

```{r}
#| eval: false
sigma <- empirical.prec(manID.UGGA, "CH4dry_ppb")$prec

# write the empirical precision into the prec column goFlux() reads ...
manID.UGGA$CH4_prec <- sigma
# ... and ask for a 95 % MDF
CH4_flux <- goFlux(manID.UGGA, "CH4dry_ppb", conf = 0.95)
CH4_best <- best.flux(CH4_flux)          # the MDF criterion now uses the 95 % MDF

# or scale an existing flux table before the MDF criterion is applied
CH4_best <- best.flux(CH4_flux, conf = 0.95)   # adds the column MDF.conf
```

::: callout-tip
#### With a stock goFlux

Nothing above is needed to get the same MDF with an older goFlux: supply $1.96 \cdot \sigma$ as `prec`. Either at import (`import.UGGA(..., prec = 1.96 * sigma)`) or directly in `goFlux(..., prec = 1.96 * sigma)`; the `MDF` column is then the 95 % limit.
:::

## Reporting

State the three ingredients: the precision estimator (datasheet, closure-median second-difference σ or per-measurement Allan deviation), the confidence level, and the measurement time in seconds, e.g. "MDF at 95 % (two-sided), second-difference precision of 2.2 ppb (median of 14 closures, 1 s logging), 300 s closure". The flux table carries `prec`, `MDF` and, when `best.flux(conf = )` was used, `MDF.conf`.

::: {.callout-note}
The p-value of the linear slope (`LM.p.val`) is not a detection criterion: on 1 Hz records the residuals are autocorrelated and the ordinary least-squares test is anticonservative. Use it as a coarse validity filter and the MDF with an empirical precision and a stated `conf` for detection.
:::
